Xolotl: An Intuitive and Approachable Neuron and Network Simulator for Research and Teaching

### 2018-10-26 paper/paper.tex
Fix path to mechanisms

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like Python or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009, gratiy2018bionet}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), ANNarchy (\cite{vitayANNarchyCodeGeneration2015}), BRIAN (\cite{stimbergBrianSecondComing2013}), morphforge (\cite{hullMorphforgeToolboxSimulating2014}), and PyNN (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as C or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Because models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

\paragraph{A rich library of network components.} Xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. ``Compartments'' represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components; and can represent either entire cells or parts of cells. Objects of type ``conductance'' represent populations of ion channels in a compartment that produce transmembrane currents. ``synapse'' objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type ``mechanism'' can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them. Parameters of any of these objects can be easily inspected and modified at any time, either manually or through a programmatic interface. \hl{Every compartment exists as a C++ file within the code repository, making it easy to browse these components and edit them. For example, mechanisms are stored under ``c++/mechanism''. }

@@

Network models in xolotl consist of compartment objects that can be connected by synapse objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, in the well-studied pyloric circuit in the crustacean stomatograstic ganglion, a cell called LP synapses onto PY cells using several synapses, that together have an effective combined maximal conductance of around 30 nS. This can be modelled in xolotl by using single compartments to represent the LP and PY cells, and connecting the LP cell to the PY cell using a chemical synapse of type "Glutamatergic" with strength 30 nS using

@@

Conductance-based neuron models are typically high dimensional and contain many parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: switching between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

A common theme in the architecture of many simulators and high-performance software is the union of user-facing code written in a scientific programming or user-friendly language and performance-critical code written in a fast, compiled language like C++. The reasons for splitting the codebase across two languages are often pragmatic: user interfaces and second-order tools are often easier to implement in richer programming environments, but performance requirements often mean that code needs to be written in a language like C++ or FORTRAN. While newer projects like Julia aim to merge ease of use and performance in a single language, a number of simulators and software projects embrace this division across languages.  For example, DynaSim, a general purpose dynamical systems simulator, allows for models specified in MATLAB to be compiled into C and executed, speeding up execution (\cite{sherfeyDynaSimMATLABToolbox2018}). NEURON similarly presents a more use-friendly Python interface, while integration is carried out in compiled code (\cite{hinesNEURONSimulationEnvironment1997}). MATLAB's built in "codegen" command allows MATLAB functions to be automatically translated into C code and compiled, which can then be run instead of the MATLAB version. The SWIG project (\cite{beazley1996swig}) offers tools to connect programs written in C or C++ to a variety of scripting languages like Python and Ruby.

@@

Our approach in developing "cpplab", our bridge between C++ code and MATLAB, was conceptually different from many of these architectures. Unlike MATLAB's "codegen" function, which assumes the primacy of MATLAB code and makes C code that mirrors the MATLAB code, "cpplab" places C++ code first, and only aims to mimic the structure of objects specified in C++ header files, but not their functionality. Thus, cpplab is able to create a bonafide MATLAB object whose properties reflect that of the C++ object, but not its methods. This simpler task can be easily achieved without the construction of an intermediate interface file, as used in more powerful solutions like SWIG. For the purposes of a neuron and network simulator, this architecture is more robust since only simple parsing of the C++ code is needed. This architecture can be easily used in other applications, and it is trivial to remove some restrictions that the cpplab architecture currently imposes, like lack of support for vectors.

@@

Over the years, several simulators have been developed to integrate systems of coupled differential equations that model the spiking activity of neurons (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003, hinesNEURONSimulationEnvironment1997, bower2003genesis}). A critical architectural choice in designing a simulator is how much ``scaffolding'' a user is provided with to construct a model; whether a model is specified by equations or by components, or by some combination of the two. In an equation-oriented architecture, the user starts with a blank slate and the primary method of specifying a model is to write out its differential equations (\cite{stimberg2014equation}). In contrast, in a component-oriented architecture, the primary method of specifying a model is to assemble it from pre-existing components, each of which include differential equations, parameters, and solvers. Both approaches have advantages and disadvantages that are discussed below.

An equation-oriented simulator can be more transparent and allows the user to know exactly what is being solved, but equations can be cumbersome to write out, read, or to debug. In most commonly used programming environments, these equations have to be entered as strings, and complex parsing has to be carried out by the simulator to check that these strings constitute valid equations. In addition, parameters have to be written explicitly into equations, and it is usually not trivial to change parameters after initial specification. In contrast, components are easy to assemble into a model, but it can be hard to know what they contain, where they are physically located on the user's computer, and how they can be changed. NEURON is primarily a component-oriented neuronal simulator, and new components are specified in special model files that can be ``inserted'' into a model. BRIAN is an equation-oriented neuronal simulator meant to be used from within Python (\cite{goodmanBrianSimulator2009}). XPP is a general purpose dynamical system simulator that is equation-oriented (\cite{ermentrout2002simulating}). DynaSim is an equation-oriented simulator with some component-oriented capabilities. Models can be specified by both strings of equations or components, but since models do not exist as objects in the workspace, parameters and variables have to be reinitialized when changed (\cite{sherfeyDynaSimMATLABToolbox2018}).

@@

Another architectural choice in designing simulators is the syntax required for specifying models. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units in NEURON, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulators can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load on users using more than one simulator. Here, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of the simulator.

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]
Finally, we walk through a typical workflow to compare xolotl to other neuron simulators. The first step is to initialize a xolotl object, e.g., using "x = xolotl". This is similar to importing the NEURON package from within Python. Compartments can be created and added to the object tree using "x.add(...)", and channels and mechanisms can be added to these compartments using further "add" commands. This workflow is similar to NEURON, where conductances are ``inserted'' into compartments. In both xolotl and NEURON, the tree of objects can be inspected and individual properties like maximal conductances or reversal potentials can be changed from the command line. The setup phase of the workflow is thus similar to that in NEURON, but differs from equation-oriented environments like Brian or DynaSim, where a set of equations has to be explicitly written out. xolotl differs from NEURON at this stage in that it is possible to ``click through'' the object tree all the way to view the underling C++ code for a particular component, and thus to view the ODE being solved, and the numerical method used to solve it. Since NEURON uses several intermediate files, that may have to be compiled, it is not trivial to identify and view the file that matches an object in the workspace. Another unique feature of xolotl is that every xolotl object is automatically and silently hashed in the background, and that this hash can be inspected at any time. This makes it easy to verify that the object constructed so far is what it should be.
\end{tcolorbox}

>>>>

Two major approaches have dominated the design of neuron simulators. One approach is to write the simulator in a fast compiled language like C and allow for the construction and simulation of neuron models using object-oriented paradigms. This approach has been implemented in NEURON (\cite{hinesNEURONSimulationEnvironment1997}). Simulators designed in this way tend to perform fast computations with little overhead, but suffer from a steep learning curve. Wrapping these simulators in a more approachable language like Python or using graphical user interfaces (GUIs) mitigates these drawbacks (\cite{hinesNEURONPython2009, gratiyBioNetPythonInterface2018}) at the cost of obfuscating the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, simulators designed from the ground up in popular scientific computing languages can be easier to use and benefit from interoperability with other commonly-used tools. Simulators like DynaSim (\cite{sherfeyDynaSimMATLABToolbox2018}), ANNarchy (\cite{vitayANNarchyCodeGeneration2015}), BRIAN (\cite{stimbergBrianSecondComing2013}), morphforge (\cite{hullMorphforgeToolboxSimulating2014}), and PyNN (\cite{davisonPyNNCommonInterface2009}) allow the user to specify models with strings of equations or components that are constructed using a special syntax, that can then be translated into a faster implementation language such as C or C++ (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits considerable flexibility for simulating systems of differential equations. Because models need to be translated between the two languages, the hierarchical nature of neuron models is not naturally encapsulated by these tools, and the syntax can be verbose. Neither approach facilitates the creation of tools that simultaneously maintain efficiency, ease-of-use, and clarity.

@@

\paragraph{A rich library of network components.} Xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. ``Compartments'' represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components; and can represent either entire cells or parts of cells. Objects of type ``conductance'' represent populations of ion channels in a compartment that produce transmembrane currents. ``synapse'' objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type ``mechanism'' can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them. Parameters of any of these objects can be easily inspected and modified at any time, either manually or through a programmatic interface. \hl{Every compartment exists as a C++ file within the code repository, making it easy to browse these components and edit them. For example, mechanisms are stored under "xolotl/c++/mechanisms". }

@@

Network models in xolotl consist of compartment objects that can be connected by synapse objects. Two compartments representing different neurons can be connected using synapses using the built-in "connect" method. For example, in the well-studied pyloric circuit in the crustacean stomatograstic ganglion \cite{marderUnderstandingCircuitDynamics2007}, a cell called LP synapses onto PY cells using many synapses, that together have an effective combined maximal conductance of around 30 nS. This can be modeled in xolotl by using single compartments to represent the LP and PY cells, and connecting the LP cell to the PY cell using a chemical synapse of type "Glutamatergic" with strength 30 nS using

@@

Conductance-based neuron models are typically high dimensional and contain many parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowaschFailureAveragingConstruction2002}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: switching between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

A common theme in the architecture of many simulators and high-performance software is the union of user-facing code written in a scientific programming or user-friendly language and performance-critical code written in a fast, compiled language like C++. The reasons for splitting the codebase across two languages are often pragmatic: user interfaces and second-order tools are often easier to implement in richer programming environments, but performance requirements often mean that code needs to be written in a language like C++ or FORTRAN. While newer projects like Julia (\cite{bezansonJuliaFreshApproach2017}) aim to merge ease of use and performance in a single language, a number of simulators and software projects embrace this division across languages.  For example, DynaSim, a general purpose dynamical systems simulator, allows for models specified in MATLAB to be compiled into C and executed, speeding up execution (\cite{sherfeyDynaSimMATLABToolbox2018}). NEURON similarly presents a more user-friendly Python interface, while integration is carried out in compiled code (\cite{hinesNEURONSimulationEnvironment1997}). MATLAB's built in "codegen" command allows MATLAB functions to be automatically translated into C code and compiled, which can then be run instead of the MATLAB version. The SWIG project (\cite{beazleySWIGEasyUse1996}) offers tools to connect programs written in C or C++ to a variety of scripting languages like Python and Ruby.

@@

Our approach in developing "cpplab", our bridge between C++ code and MATLAB, was conceptually different from many of these architectures. Unlike MATLAB's "codegen" function, which assumes the primacy of MATLAB code and makes C code that mirrors the MATLAB code, "cpplab" places C++ code first, and only aims to mimic the structure of objects specified in C++ header files, but not their functionality. Thus, cpplab is able to create a bonafide MATLAB object whose properties reflect that of the C++ object, but not its methods. This task can be easily achieved without the construction of an intermediate interface file, as used in more powerful solutions like SWIG. For the purposes of a neuron and network simulator, this architecture is more robust since only simple parsing of the C++ code is needed. This architecture can be easily used in other applications, and it is trivial to remove some restrictions that the cpplab architecture currently imposes, such as lack of support for vectors.

@@

Over the years, several simulators have been developed to integrate systems of coupled differential equations that model the spiking activity of neurons (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003, hinesNEURONSimulationEnvironment1997,bowerBookGENESIS2nd1998}). A critical architectural choice in designing a simulator is how much ``scaffolding'' a user is provided with to construct a model; whether a model is specified by equations or by components, or by some combination of the two. In an equation-oriented architecture, the user starts with a blank slate and the primary method of specifying a model is to write out its differential equations (\cite{stimbergEquationorientedSpecificationNeural2014}). In contrast, in a component-oriented architecture, the primary method of specifying a model is to assemble it from pre-existing components, each of which include differential equations, parameters, and solvers. Both approaches have advantages and disadvantages that are discussed below.

An equation-oriented simulator can be more transparent and allows the user to know exactly what is being solved, but equations can be cumbersome to write out, read, or to debug. In most commonly used programming environments, these equations have to be entered as strings, and complex parsing has to be carried out by the simulator to check that these strings constitute valid equations. In addition, parameters have to be written explicitly into equations, and it is usually not trivial to change parameters after initial specification. In contrast, components are easy to assemble into a model, but it can be hard to know what they contain, where they are physically located on the user's computer, and how they can be changed. NEURON is primarily a component-oriented neuronal simulator, and new components are specified in special model files that can be ``inserted'' into a model. BRIAN is an equation-oriented neuronal simulator meant to be used from within Python (\cite{goodmanBrianSimulator2009}). XPP is a general purpose dynamical system simulator that is equation-oriented (\cite{ermentroutSimulatingAnalyzingAnimating2002}). DynaSim is an equation-oriented simulator with some component-oriented capabilities. Models can be specified by both strings of equations or components, but since models do not exist as objects in the workspace, parameters and variables have to be reinitialized when changed (\cite{sherfeyDynaSimMATLABToolbox2018}).

@@

Another architectural choice in designing simulators is the syntax required for specifying models. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulators can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load on users working with more than one simulator. Here, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of the simulator.

\begin{tcolorbox}[breakable,notitle,boxrule=0pt,colback=yellow,colframe=yellow]
Finally, we walk through a typical workflow to compare xolotl to other neuron simulators. First, initialize a xolotl object (e.g. "x = xolotl"). This is similar to importing the NEURON or Brian packages from within Python. Compartments can be created and added to the object tree using "x.add(...)", and channels and mechanisms can be added to these compartments using further "add" commands. This workflow is similar to NEURON, where conductances are ``inserted'' into compartments. In both xolotl and NEURON, the tree of objects can be inspected and individual properties like maximal conductances or reversal potentials can be changed from the command line. The setup phase of the workflow is therefore similar to that in NEURON, but differs from equation-oriented environments like Brian or DynaSim, where a set of equations has to be explicitly written out. Unlike in NEURON, it is possible to ``click through'' the xolotl object tree all the way down to view the underling C++ code for a particular component, and thus to view the ODE being solved, and the numerical method used to solve it. Since NEURON uses several intermediate files, that may have to be compiled, it is not trivial to identify and view the file that matches an object in the workspace. While NEURON requires compilation of a custom binary version including all network components, xolotl automatically and silently hashes the network in the background to determine which files must be compiled and linked. This hash can be inspected at any time, making it easy to verify that the object constructed so far is what it should be.
\end{tcolorbox}
